---
title: "BBS Trend Summaries for Grassland Birds"
output:
  beamer_presentation: default
  ioslides_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

BBS Case Study
=========

- Data from the North American Breeding Bird Survey (BBS)
- Goal of the yearly survey is to monitor bird population at large geographic scales
- Data collected for over 400  species (focus on 28 grassland species)
- Interested in measuring trend from 1966 to 1999

The Data
=========

- Focus on summaries of temporal trend parameters $\beta_i$ for species $1, ..., K$.
- For each of 28 species, have trend estimate $\hat \beta_i$, the corresponding standard error $SE_i$, and the number of sites $n_i$.


The Model
===============

Sampling Model:

1.  Assume the estimates $\hat \beta_i$ are independent

2.  ${\hat \beta}_i \sim N(\beta_i, 1 / \tau({\hat \beta}_i))$

3.  $SE_i \sim Gamma(n/2, (n/2) \tau({\hat \beta}_i))$


Combining Estimates?
============

- No-pool model
- Pool model
- Multilevel (partially pool) model

Read in the Data
======

```{r}
library(tidyverse)
(d <- read_csv("BBS_survey.csv"))
```

No-Pool Model
========

- No relationship between $\beta_1, ..., \beta_N$

- Estimate $\beta_j$ with the estimate $\hat \beta_j$

- Use JAGS to construct interval estimates

Prior
==========

- $\beta_i \sim N(0, 0.001)$
- $\tau({\hat \beta}_i) \sim G(0.001, 0.001)$

JAGS model script
=============

```{r}
model_string <- "model{
  for (s in 1:28) {
      varhat[s] ~ dgamma(p[s], lam[s])
      p[s] <- n[s] / 2
      lam[s] <- p[s] * tau.betahat[s]
      tau.betahat[s] ~ dgamma(0.001, 0.001)
      sd.betahat[s] <- 1 / sqrt(tau.betahat[s])
      betahat[s] ~ dnorm(beta[s], tau.betahat[s])
      beta[s] ~ dnorm(0, 0.0001)
  }
}"
```

JAGS Data
==========

```{r}
betahat <- d$Trend
varhat <- d$SE ^ 2
n <- d$N_Site
the_data <- list(betahat = betahat,
                 varhat = varhat,
                 n = n)
```

JAGS Run
=============

```{r}
library(runjags)
posterior <- run.jags(model_string,
                      n.chains = 1,
                      data = the_data,
                      monitor = c("beta"),
                      adapt = 1000,
                      burnin = 10000,
                      sample = 5000)
```


Posterior Interval Estimates
=============

```{r}
library(coda)
library(bayesplot)
posterior %>% as.mcmc() %>% as.data.frame() -> post
select(post, `beta[1]`:`beta[28]`) %>%
  mcmc_intervals() 
```

Comments
============

- Some of the individual estimates are relatively poor
- Desirable to strengthen these "small n" estimates by borrowing information from related species

Pooled Model
===========

- Suppose the trends are the same across species

- Assume $\beta_1 = ... = \beta_N = \beta$

- Assign $\beta$ and $\tau$ parameters vague priors

- Revise JAGS script

JAGS Script
=========

```{r}
model_string <- "model{
  for (s in 1:28) {
      varhat[s] ~ dgamma(p[s], lam[s])
      p[s] <- n[s] / 2
      lam[s] <- p[s] * tau.betahat[s]
      tau.betahat[s] ~ dgamma(0.001, 0.001)
      sd.betahat[s] <- 1 / sqrt(tau.betahat[s])
      betahat[s] ~ dnorm(beta, tau.betahat[s])
  }
  beta ~ dnorm(0, 0.0001)
}"
```

JAGS Run
===========

```{r}
library(runjags)
posterior <- run.jags(model_string,
                      n.chains = 1,
                      data = the_data,
                      monitor = c("beta"),
                      adapt = 1000,
                      burnin = 10000,
                      sample = 5000)
```

Summary of $\beta$
=========

```{r}
summary(posterior) 
```

Comments
============

- Pooled estimate only is reasonable if know that species all have the same trend

- Lose information about individual species

- A compromise estimate is desirable

The Prior (Multilevel)
===============

1. Assume $\beta_1, ..., \beta_{N}$ are iid $N(\mu, 1 /\tau)$

2.  Assign $(\mu, \tau)$ a vague prior


Fitting the Model using JAGS
===============

Write a script defining the model.

```{r}
model_string <- "model{
  for (s in 1:N) {
      varhat[s] ~ dgamma(p[s], lam[s])
      p[s] <- n[s] / 2
      lam[s] <- p[s] * tau.betahat[s]
      tau.betahat[s] ~ dgamma(0.001, 0.001)
      sd.betahat[s] <- 1 / sqrt(tau.betahat[s])
      betahat[s] ~ dnorm(beta[s], tau.betahat[s])
      beta[s] ~ dnorm(mu, tau.beta)
  }
  mu ~ dnorm(0.0, 1.0E-6)
  tau.beta ~ dgamma(0.001, 0.001)
  sd.beta <- 1 / sqrt(tau.beta)
  another.beta ~ dnorm(mu, tau.beta)
}"
```

Define the data.
=============

```{r}
d <- read_csv("BBS_survey.csv")
betahat <- d$Trend
varhat <- d$SE ^ 2
n <- d$N_Site
the_data <- list(betahat = betahat,
                 varhat = varhat,
                 n = n,
                 N = length(n))
```

Compile and run the model.
=============

```{r}
library(runjags)
posterior <- run.jags(model_string,
                      n.chains = 1,
                      data = the_data,
                      monitor = c("beta", "mu", "sd.beta"),
                      adapt = 1000,
                      burnin = 20000,
                      sample = 5000)
```

Running JAGS
================

Create a matrix of simulated draws.

```{r}
library(coda)
posterior %>% as.mcmc() %>% as.data.frame() -> post
```

Interval Estimates
==================

```{r}
select(post, `beta[1]`:`beta[28]`) %>%
  mcmc_intervals()
```

Derived Parameters of Interest
==========

- How many of these species have negative trends? 

- That is, we are interested in the derived parameter
$$
D(\beta_1, ..., \beta_{28}) = \sum_{i=1}^{28} I(\beta_i < 0)
$$

Derived Parameters of Interest
==========

- What is the actual rank of the 27th species?

$$
R_{27} = \sum_{i=1}^{28} I(\beta_{27} \ge \beta_i)
$$

Easy to Learn About Derived Parameters
=========

- Have matrix of simulated draws from the posterior of $\beta_1, ..., \beta_N$

- For each row, compute derived parameter $D(\beta)$.

- Have simulated sample from posterior of $D(\beta)$

Posterior of D
=================

```{r, echo = FALSE}
library(TeachBayes)
```

```{r, fig.height = 5.5}
Beta <- post[, 1:28]
sum.negative <- function(y) {sum(y < 0)}
D <- apply(Beta, 1, sum.negative)
bar_plot(D)
```

Posterior of rank of R_27 ?
===============

```{r, fig.height = 5.5}
Ranks <- t(apply(Beta, 1, rank))
bar_plot(Ranks[, 27]) + xlim(0, 13)
```


